\textsf{\textbf{Drop Electro-Bouncing in Low-Gravity}}

### 2018-03-22 doc/manuscript/manuscript.tex
added presentation latex scratch pad
fotmatting changes, updated figures

\documentclass[12pt,a4paper,oneside]{book}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{fontspec} % This line only for XeLaTeX and LuaLaTeX
\usepackage{pgfplots}
\usepackage{pgf}
\graphicspath{ {../figures/} }
%\usepackage{svg}
\usepackage{verbatim}
\usepackage{color,soul}
\usepackage{listings}
\usepackage{setspace}
\newlength\figureheight
\newlength\figurewidth
\setlength\figureheight{7cm}
\setlength\figurewidth{10cm}
\usepackage[
backend=biber,
style=phys,
sorting=none
]{biblatex}
\addbibresource{thesis.bib}
\usepackage{import}
\let\pgfimageWithoutPath\pgfimage 
\renewcommand{\pgfimage}[2][]{\pgfimageWithoutPath[#1]{../figures/#2}}
\usepackage{array} 
\usepackage{appendix}
\usepackage{makeidx}
\makeindex

@@

\begin{document}
%\frontmatter
%\maketitle
%\afterpage{\blankpage}
\begin{titlepage}
\doublespacing

@@

\vspace{12ex}
Thesis Committee:\\
Mark Weislogel, Chair\\
Ra\'ul Bayo\'an Cal\\
Gerry Recktenwald\\

@@

\doublespacing
\clearpage

@@

\doublespacing
\section*{Abstract}
\addcontentsline{toc}{chapter}{\numberline{Abstract}}
\noindent
We investigate the dynamics of spontaneous jumps of water droplets from electrically charged superhydrophobic dielectric substrates during a sudden step reduction in gravity level. In the brief free-fall environment of a drop tower, with a non-homogeneous external electric field (with strengths ~0.39-2.36 kV/cm) arising due to dielectric surface charges, body forces acting on the jumped droplets are primarily supplied by polarization stress and Coulombic attraction instead of gravity. This electric body force leads to a droplet bouncing behavior similar to well-known phenomena in 1-$g_0$, though occurring for much larger droplets ($\sim$ 0.5 mL). We show a simple model for the phenomenon, its scaling, and asymptotic estimates for droplet time of flight. The droplet trajectories are controlled primarily by the dimensionless electrostatic Euler number $\mathbb{E}\mbox{u}$, which is a ratio of inertial to electrostatic forces. For ${\mathbb{E}\mbox{u}}_+ \gg 1/8 \pi$ droplets escape the electric field. The droplet net charge, estimated to be on the order of $1 \times 10^{-9}$ C, is field-induced rather than by contact charging at the (PTFE coated) hydrophobic interface. In 1-$g_0$, for Weber numbers $>$ 0.4, impact recoil behavior on a super-hydrophobic surface is normally dominated by damping from contact line hysteresis. However, for the low Bond and Ohnesorge numbers occurring in free-fall, the droplet impact dynamics additionally include electrohydrodynamic surface wettability effects. This is qualitatively discussed in terms of trends in coefficients of restitution and dimensionless contact time.

@@

\mainmatter
\chapter{Introduction}
\begin{quote}
``When the influence of gravity on fluid behavior is diminished or removed, other forces, otherwise of small significance, can assume paramount roles.''
- NRC Report to NASA, 2003  \cite{motil_priorities_2012}
\end{quote}

Our terrestrially born intuition about how liquids flow is easily confounded in a low-gravity environment. This should come as little surprise, as we are creatures evolved at the bottom of a gravity well; gravity is natural to us. However when the magnitude of gravitational body forces become small, other forces come into play in fluid dynamical problems which are otherwise negligible under normal circumstances in 1-$g_0$. Our 1-$g_0$ cognitive bias then leads to a plethora of problems which are both simultaneously trivially easy to solve in 1-$g_0$, and which despite decades of study continue to elude solutions in a low-gravity context. One example of these problems is the so-called ``phase separation'' problem of separating a gas phase from a multi-phase flow (or the reverse case), without the aid of buoyancy due to gravity. Sans buoyancy, otherwise mundane activities such as venting a gas or settling a liquid in a tank become problematic \cite{petrash_controlling_1964}. Trapped bubbles can vapor lock life support, power, or propulsion systems \cite{jenson_passive_2014}. Issues of phase separation have so bedeviled human endeavors in space that an entire Apollo Saturn 1B mission (AS-203) was earmarked to study them \cite{hastings_[saturn_1965}. This problem has for some time motivated a quest for substitute body forces, and the present work follows in that well-worn tradition.

A solid beginning to rectifying our problem of poor low-gravity intuition, would be to quantify how wrong our intuition is. First, let's clear up a common misnomer about the state of ``zero-gravity''. The theory of general relativity is based on a fundamental equivalence of gravity and acceleration. By measuring acceleration alone, one cannot determine whether one is in gravity-free space or free-falling in a gravitational field (or conversely whether or not one is uniformly accelerating, or standing at rest in a gravitational field). This is sometimes called the equivalence principle of general relativity (e.g. the equivalence of gravitational and inertial mass). In theoretical terms, a zero-g experiment performed in a drop tower on the surface of the Earth is equivalent to the same experiment performed in the nearly gravity-free deep space between galaxies, despite the fact that the drop tower itself is enveloped in a 1-$g_0$ gravity field. At the orbital altitude of the International Space Station (ISS) the gravity level is still nearly 90\% of its value at sea-level on Earth, but the astronauts aboard ISS feel weightlessness because they are in free-fall. In a practical sense true zero-g does not exit anywhere in the universe as there are always small, but nonetheless, measurable forces which can accelerate any massive body. For instance in low-Earth orbit, accelerations due to aerodynamic resistance in the tenuous outer atmosphere are on the order of $1 \times 10^{-6}$ g (from hence comes the term ``microgravity''). This small drag acceleration provides some ersatz gravity, but there are also other forces present. Foremost of these are capillary forces.

@@

As with capillary forces, electrostatic forces have a multitude of applications as substitutes for gravity. Electrohydrodynamic (EHD) body forces have been suggested to combat thermal stratification of propellants under low-g conditions where bouyancy induced natural convection is very small \cite{blackmon_collection_1965}. In this case electro-convection is an analog of natural convection due to a fluid temperature dependent dielectric permitivity gradient (as opposed to a fluid density gradient), in the presence of a body force field: the electric field. The study also suggested applications in cryogenic tank vent screening, and in reducing heat transfer to cryogenic propellants (and thus boiloff losses) by dielectrophoretically positioning ullage vapors around the tank walls as a thermal barrier. EHD forces have been studied for enhancing boiling heat transfer by promotion of bubble detachment and prevention of dryout, again as an ersatz bouyancy \cite{snyder_dielectrophoresis_2001} \cite{di_marco_influence_2003} \cite{marco_use_2012}. EHD heat pipes, which substitute an electrode structure for the capillary wicking structure of a conventional thermocapillary heat pipe can evade the wicking limit \cite{jones_electrohydrodynamic_1973}. These are restricted to the use of insulating dielectric liquids, which usually have relatively low thermal conductivity, but are highly robust against dryout, support active flow control, and have low viscous losses. Dielectrophoretic settling of cryogenic propellants, in both total and partial communication PMD configurations, has been studied analytically in both static \cite{hurwitz_electrohydrodynamic_1966} and dynamic cases \cite{koval_dynamics_1967}, and in drop tower experiments \cite{fax_dielectrophoretic_1969}. EHD forces have also beeen applied to the low-g slosh baffling problem \cite{boretz_orbital_1970} \cite{petrash_use_1968} \cite{hurwitz_dielectrophoretic_1968}, and studied extensively for the mitigation of vapor pullthrough during tank draining, and concomittent minimization of propellant residuals at burnout, in a series of experiments at the NASA Lewis (now Glenn) Research Center 2.2 s drop tower facillity \cite{berenyi_dielectrophoretic_1970}.

\section{Spontaneous Drop Jump}
Low-gravity phase separation, droplet dynamics, and the dynamics and applications of the so-called spontaneous droplet jump, is an active area of study at the Portland State University Dryen Drop Tower labratory (DDT). When a nonwetting, gravity dominated sessile droplet (e.g. a puddle), which is initially at rest on a surface in the Cassie-Baxter state suddenly undergoes a large step reduction in $\mathbb{B}\mbox{o}$ it will spontaneously jump away from the surface. The Cassie-Baxter state, sometimes also known as the Fakir state, is a metastable heterogenously non-wetting state on a textured surface characterized by the existance of a vapor-layer which seperates roughness features of the surface from the bottom of the droplet. The spontaneous drop jump was first observed experimentally in the Soviet Union by Kirko \emph{et al.} \cite{kirko_phenomenon_1970} in 1970 for drops of mercury in HCl, and later by Wollman \emph{et al.} in 2016 for water drops in a set of experiments conducted using a low-gravity drop tower \cite{wollman_more_2016}. The kinetic energy of the jump is supplied by the defect in free surface energy as the new minimum energy surface equilibrium has approximately constant curvature. If the energy losses by viscous and contact line dynamics are neglected, the available kinetic energy is given by
\[KE_2 = SE_1 - SE_2= [(\sigma A)_{ls} + (\sigma A)_{gl} + (\sigma A)_{sg}]_1 - [(\sigma A)_{gl} + (\sigma A)_{sg}]_2, \]

where the subscripts $_1$ and $_2$ denote the intial 1-$g_0$, and low-g surface energies. The motive force is due to the inertia of internal flows which occur as the interface deflects under the suddenely lessened $\mathbb{B}\mbox{o}$. \hl{The characteristic time scale of the rolling up scales as} $t_j \sim R_p/U ~(\rho H R^2_p/\sigma)^{1/2}$ \cite{attari_puddle_2016}, which resembles the contact time, $\tau \approx 2.6(\rho R^3_d/\sigma)^{1/2}$ presented by Richard \emph{et al.} in 2002 for the related problem of drops impacting hydrophobic surfaces in 1-$g_0$ \cite{richard_surface_2002}. For drops with radial symmetry and sufficiently high initial $\mathbb{B}\mbox{o}$ inward radial capillary waves coalesce at the axis, leading to geysering and creation of satellite droplets by the Rayleigh-Plateau breakup of the geyser. In the case of smaller jumping drops the capillary waves are gradually damped by viscous forces. A time-lapsed composite photograph of spontaneous drop jump can be seen in Figure \ref{fig:nobounce}
\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{drop_jump}
\caption{The trajectory of a 0.2 mL drop is captured in a composite image over a period $\sim 0.67$ s at 120 fps. \label{fig:nobounce}}
\end{figure}

@@

\section{Drop Electro-Bounce}
During the `rolling-up' of drops under ideal conditions, the spontaneous jump phenomenon is governed by a balance of inertia and surface tension forces, and once aloft the drop motion is nominally in a regime of pure drag, however other forces can again come into play.

We have observed jumped drops to occasionally decelerate and return to the superhydrophobic surface, rebounding multiple times in the fashion of rigid bodies bouncing under 1-$g_0$. The forces at work in such situations are presumably electrostatic in origin. Though this was an unintended effect in our work on spontaneous drop jump, but the phenomenon is interesting itself. There is a certain irony to this typical failure of low-gravity intuition (the effect of relatively tiny electrostatic forces becoming suddenly large in the absence of gravity), being as we are experienced low-gravity researchers who perhaps should know better. A time-lapsed composite image showing an example of the phenomenon is shown in Figure \ref{fig:bounce}.
\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{bounce}
\caption{The trajectory of a 0.5 mL drop is captured in a composite image over a single bounce period ($\sim 1.25$ s). The surface potential of the superhydrophobic dielectric in this example is $\varphi_s = 1.25 \pm 0.41$ kV. \label{fig:bounce}}
\end{figure}

@@

There are several possible origins of the electric charge which are ultimately responsible for this phenonenon. It is well known that water acquires positive free charge when in contact with certain polymers, especially polytetrafluoroethylene (PTFE) \cite{langmuir_surface_1938}, through a process called contact charging. PTFE, on the other hand, tends to readily acquire negative chage by contact with water. The superhydrophobic surfaces used in the spontanous drop jump experiments have thin (nanometric) PTFE coatings, and we observe that it is extremely easy to produce significant surface potentials ($\varphi_s \sim$ 100-500 V) by simply flowing streams of distilled water over them. A study of this water on PTFE contact charging phenomenon was conducted by Yatsuzuka \emph{et al.} \cite{yatsuzuka_electrification_1994}, who suggest that this process results from formation of an electrical double layer driven by selective adsorption of ($\mbox{OH}^-$) ions at the polymer surface; other recent work strongly supports this hypothesis \cite{beattie_intrinsic_2006} \cite{strazdaite_water_2015}. Yatsuzuka \emph{et al.} also found that the specific charge on the drops in contact with the surface depends on their both surface velocity and conductivity. The contact charging mechanism is the most likely accidental source of charge on the suphydrophobic surfaces used in our work. Given the large roughness, or the ratio of projected to actual surface area, of the superhydrophobic surfaces used in the experiment, and given that the drops are initially in a Cassie-Baxter state, a somewhat electrically resistive air layer is maintained that reduces grounding of the drops despite the large potential difference between them and and the surface charges. If this electrostatic potential is greater than the release of surface energy under the sudden change in $\mathbb{B}\mbox{o}$ (in the form of the drop kinetic energy), we expect that the drop will return to an equilibrium state on the charged surface which minimizing the potential.

The source of the net free charge on the drops is another issue. The drop charge could of course also be due to the contact charging mechanism mentioned previously. For instance, in a 1996 paper, NASA flight engineer Don Pettit discusses the problem of low-gravity flow induced charging of liquids, resulting ultimately from contact charging phenomenon \cite{pettit_donald_flow_????}. Another mechanism for the drop charge is field-induced charging. Field-induced charging occurs due to physical breakup of a conductor having a field-induced dipole (e.g a physical seperation of charge). In our case this occurs when the drop is deposited on the charged surface by a syringe. The metal syringe needle tip, and the liquid in the syrigne itself are essentially a ground connection which is broken when the syringe is removed. Field-induced charging is at work in the famous Kevin thunderstorm, and is applied in inkjet, and electrospay technologies, where in each case the breakup is by the Rayleigh-Plateau instability. Notably, in Pettit's discussion of contact charging of liquids in low-g, he remarks on accidental electrostatic `hula-ing' of silicone oil drops when ejected from a syringe in the vicinity of a highly-charged polymer surface during an experiment conducted aboard STS-5 by mission specialist Joseph Allen \cite{pettit_donald_flow_????}. Depending on the (highly-variable) electrical conductivity of the silicone oil, and the material of the syringe used in the experiment, the charge could arise just as easily by field-induction as by contact charging. Relatedly, in a series of informal, and somewhat whimsical experiments, Pettit himself electrostatically orbited small water drops around a triboelectrically charged PTFE knitting needle while aboard ISS during expedition 30/31 \cite{stevenson_electrostatic_2015}. Again, in this case, the drop charge is likely field-induced.

\section{Applications}
This study of the drop electro-bounce phenomenon may have results which can be extended to useful applications in a space environment. External surfaces of spacecraft tend to become charged with time due to the space environment itself. Surface charging largely occurs due to low energy electrons (3-50 keV) which do not penetrate the surface of the spacecraft external stucture \cite{czepiela_charging_1997}. The ultimate sources of these charges are trapped charged particles of the van Allen radiation belts, galactic cosmic rays, and the solar wind. This deposited charge accumulates and can lead to significant potential differences between different parts of a spacecraft, sometime leading to breakdown and discharge, called Paschen discharge. Deep dielectric charging occurs when higher energy charged particles penetrate the surface of a dielectric material, which can also lead to large potential differences if the dielectric leakage is lower than the external charging rate. In a fluid mechanical context, this accumulated charge could be problematic during any multiphase venting process, as might be expected to occur during autonomous satellite servicing refueling operations, or during boil-off venting of cryogens stored over long periods in a space environment (as in many crewed Mars mission scenarios, or in cryogenic propellant depot operations). Such venting, if occurring in a region of strong field, will also tend to produce a stream of charged droplets, which can then return and lead to contamination of the spacecraft surfaces.

As previously noted, robust phase separation is critical to high reliability multiphase systems used in thermal control and life support. Active (but solid-state) phase seperation for life support multiphase flows, espcially high void-fraction flows is a possible application of this work. Such flows pertain to dish washing, laundry, waste solids drying, food processing, the Sabatier CO$_2$ reaction, and possibly in vapor-compression cycle condensers. Phase seperation for other disperse droplet flows include electrostatic droplet seperators for high-efficiency Rankine-cycle turbines (which require a droplet-free vapor phase entering the turbine, but where convertional superheat approaches comes with severe mass penalties) \cite{unterberg_zero_1962}, or more speculatively, in high temperature liquid droplet radiators with electrostatic collection \cite{white_liquid_1987}. Removal of satellite droplets produced during pippetting in wet-lab research outside of a glovebox environment aboard ISS has also been recently suggested as a application of the work \hl{[ref, personal discussion with NASA MSFC folk at ASGSR17 conference]}. Drops can become spontaneously charged by contact with standard micropipette tips \cite{choi_spontaneous_2013}, and this free charge can possibly be leveraged for the purposes of phase separation in low-gravity.

\section{Overview of this Thesis}
In this thesis we develop, and partly validate a simple model to aid in design of future electrostatic disperse drop phase separators. To this end the work encompasses:
\begin{enumerate}
\item A mathematical model of the drop electro-bounce, its scale analysis, asymptotic estimates for drop apoapses and times-of-flight.
\item The results of a controlled experiment, where the effect of key independent variables drop volume $V_d$, and dielectric surface charge density $\sigma$, on drop trajectories is tested.
\item Science of opportunity comprising experimental data on drop impacts at very low Weber and Ohnesorge numbers, for varying levels of electrostatic Bond number. 
\end{enumerate}

>>>>

\documentclass[12pt,a4paper,oneside]{book}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{fontspec} % This line only for XeLaTeX and LuaLaTeX
\usepackage{pgfplots}
\usepackage{pgf}
\usepackage{bookmark}
\usepackage{color}
\usepackage{tikz}
\graphicspath{ {../figures/} }
%\usepackage{svg}
\usepackage{verbatim}
\usepackage{color,soul}
\usepackage{listings}
\usepackage{setspace}
\usepackage[caption=false]{subfig}
\newlength\figureheight
\newlength\figurewidth
\setlength\figureheight{7cm}
\setlength\figurewidth{10cm}
\usepackage[
backend=biber,
style=phys,
sorting=none
]{biblatex}
\addbibresource{thesis.bib}
\usepackage{import}
\let\pgfimageWithoutPath\pgfimage 
\renewcommand{\pgfimage}[2][]{\pgfimageWithoutPath[#1]{../figures/#2}}
\usepackage{array} 
\usepackage{appendix}
\usepackage{makeidx}
\makeindex

@@

\begin{document}

\newcommand{\redline}{\raisebox{2pt}{\tikz{\draw[-,red,solid,line width = 1.5pt](0,0) -- (5mm,0);}}}
%\frontmatter
%\maketitle
%\afterpage{\blankpage}
\begin{titlepage}
\doublespacing

@@

\vspace{12ex}
Thesis Committee:\\
Mark Weislogel, Chair\\
Ra\'ul Bayo\'an Cal\\
Gerald Recktenwald\\

@@

%\doublespacing
\clearpage

@@

%\doublespacing
\section*{Abstract}
\addcontentsline{toc}{chapter}{\numberline{Abstract}}
\noindent
We investigate the dynamics of spontaneous jumps of water droplets from electrically charged superhydrophobic dielectric substrates during a sudden step reduction in gravity level. In the brief free-fall environment of a drop tower, with a non-homogeneous external electric field (with strengths ~0.39-2.36 kV/cm) arising due to dielectric surface charges, body forces acting on the jumped droplets are primarily supplied by polarization stress and Coulombic attraction instead of gravity. This electric body force leads to a droplet bouncing behavior similar to well-known phenomena in 1-$g_0$, though occurring for much larger droplets ($\sim$ 0.5 mL). We show a simple model for the phenomenon, its scaling, and asymptotic estimates for droplet time of flight. The droplet trajectories are controlled primarily by the dimensionless electrostatic Euler number $\mathbb{E}\mbox{u}$, which is a ratio of inertial to electrostatic forces. For ${\mathbb{E}\mbox{u}}_+ \gg 1/8 \pi$ droplets escape the electric field. The droplet net charge, estimated to be on the order of $1 \times 10^{-9}$ C, is field-induced rather than occurring by contact charging at the (PTFE coated) hydrophobic interface. In 1-$g_0$, for Weber numbers $>$ 0.4, impact recoil behavior on a super-hydrophobic surface is normally dominated by damping from contact line hysteresis. However, for the low Bond and Ohnesorge numbers occurring in free-fall, the droplet impact dynamics additionally include electrohydrodynamic surface wettability effects. This is qualitatively discussed in terms of trends in coefficients of restitution and dimensionless contact time.

@@

\mainmatter
\chapter{Introduction}
\begin{quote}
``When the influence of gravity on fluid behavior is diminished or removed, other forces, otherwise of small significance, can assume paramount roles.''
- NRC Report to NASA, 2003\cite{motil_priorities_2012}
\end{quote}

Our terrestrially born intuition about how liquids flow is easily confounded in a low-gravity environment. This should come as little surprise, as we are creatures evolved at the bottom of a gravity well; gravity is natural to us. However when the magnitude of gravitational body forces become small, other forces come into play in fluid dynamical problems, which are otherwise negligible under normal circumstances in 1-$g_0$. Our 1-$g_0$ cognitive bias then leads to a plethora of problems which are both simultaneously trivially easy to solve in 1-$g_0$, and which despite decades of study continue to elude solutions in a low-gravity context. One example of these problems is the so-called ``phase separation'' problem of separating a gas phase from a multi-phase flow (or the reverse case), without the aid of buoyancy due to gravity. Sans buoyancy, otherwise mundane activities such as venting a gas or settling a liquid in a tank become problematic\cite{petrash_controlling_1964}. Trapped bubbles can vapor lock life-support, power, or propulsion systems\cite{jenson_passive_2014}. Issues of phase separation have so bedeviled human endeavors in space that an entire Apollo Saturn 1B mission (AS-203) was earmarked to study them\cite{hastings_saturn_1965}. This problem has for some time motivated a quest for substitute body forces, and the present work follows in that well-worn tradition.

A solid beginning to rectifying our problem of poor low-gravity intuition, would be to quantify how wrong our intuition is. First, let us clear up a common misnomer about the state of ``zero-gravity''. The theory of general relativity is based on a fundamental equivalence of gravity and acceleration. By measuring acceleration alone, one cannot determine whether one is in gravity-free space or free-falling in a gravitational field (or conversely whether or not one is uniformly accelerating, or standing at rest in a gravitational field). This is sometimes called the equivalence principle of general relativity (e.g. the equivalence of gravitational and inertial mass). In theoretical terms, a zero-g experiment performed in a drop tower on the surface of the Earth is equivalent to the same experiment performed in the nearly gravity-free deep space between galaxies, despite the fact that the drop tower itself is enveloped in a 1-$g_0$ gravity field, where $g_0=9.81$ m/s$^2$ is the standard gravitational acceleration at sea-level . At the orbital altitude of the International Space Station (ISS) the gravity level is still nearly 90\% of its value at sea-level on Earth, but the astronauts aboard ISS feel weightlessness because they are in a state of free-fall. In a practical sense true zero-g does not exit anywhere in the universe as there are always small, but nonetheless, measurable forces which can accelerate any massive body. For instance in low-Earth orbit, accelerations due to aerodynamic resistance in the tenuous outer atmosphere are on the order of $1 \times 10^{-6}$ $g_0$ (from hence comes the term ``microgravity''). This small drag acceleration provides some ersatz gravity, but there are also other forces present. Foremost of these are capillary forces.

@@

As with capillary forces, electrostatic forces have a multitude of applications as substitutes for gravity. Electrohydrodynamic (EHD) body forces have been suggested to combat thermal stratification of propellants under low-g conditions where bouyancy induced natural convection is very small\cite{blackmon_collection_1965}. In this case electro-convection is an analog of natural convection due to a fluid temperature dependent dielectric permitivity gradient (as opposed to a fluid density gradient), in the presence of a body force field: the electric field. The study also suggested applications in cryogenic tank vent screening, and in reducing heat transfer to cryogenic propellants (and thus boiloff losses) by dielectrophoretically positioning ullage vapors around the tank walls as a thermal barrier. EHD forces have been studied for enhancing boiling heat transfer by promotion of bubble detachment and prevention of dryout, again as an ersatz bouyancy\cite{snyder_dielectrophoresis_2001}\cite{di_marco_influence_2003}\cite{marco_use_2012}. EHD heat pipes, which substitute an electrode structure for the capillary wicking structure of a conventional thermocapillary heat pipe can evade the wicking limit\cite{jones_electrohydrodynamic_1973}. These are restricted to the use of insulating dielectric liquids, which usually have relatively low thermal conductivity, but are highly robust against dryout, support active flow control, and have low viscous losses. Dielectrophoretic settling of cryogenic propellants, in both total and partial communication Propellant Management Device (PMD) configurations, has been studied analytically in both static\cite{hurwitz_electrohydrodynamic_1966} and dynamic cases\cite{koval_dynamics_1967}, and in drop tower experiments\cite{fax_dielectrophoretic_1969}. EHD forces have also been applied to the low-g slosh baffling problem\cite{boretz_orbital_1970}\cite{petrash_use_1968}\cite{hurwitz_dielectrophoretic_1968}, and studied extensively for the mitigation of vapor pullthrough during tank draining, and concomittent minimization of propellant residuals at burnout, in a series of experiments at the NASA Lewis (now Glenn) Research Center 2.2 s drop tower facillity\cite{berenyi_dielectrophoretic_1970}.

\section{Spontaneous Drop Jump}
Low-gravity phase separation, droplet dynamics, and the dynamics and applications of the so-called spontaneous droplet jump, is an active area of study at the Portland State University Dryden Drop Tower labratory (DDT). When a nonwetting, gravity dominated sessile droplet (e.g. a puddle), which is initially at rest on a surface in the Cassie-Baxter state suddenly undergoes a large step reduction in $\mathbb{B}\mbox{o}$ it will spontaneously jump away from the surface. The Cassie-Baxter state, sometimes also known as the Fakir state, is a metastable heterogenously non-wetting state on a textured surface characterized by the existance of a vapor-layer which seperates roughness features of the surface from the bottom of the droplet. The spontaneous drop jump was first observed experimentally in the Soviet Union by Kirko \emph{et al.}\cite{kirko_phenomenon_1970} in 1970 for drops of mercury in hydrochloric acid, and later by Wollman \emph{et al.} in 2016 for water drops in a set of experiments conducted using drop towers\cite{wollman_more_2016}. The kinetic energy of the jump is supplied by the defect in free surface energy as the new minimum energy surface equilibrium has approximately constant curvature. If the energy losses by viscous and contact line dynamics are neglected, the available kinetic energy is approximated by
\[KE_2 = SE_2 - SE_1 = [(\gamma A)_{gl} + (\gamma A)_{sg}]_2 - [(\gamma A)_{ls} + (\gamma A)_{gl} + (\gamma A)_{sg}]_1, \]

where $KE$ and $SE$ are the kinetic and surface energies, $A$ is the surface area, the subscripts $_1$ and $_2$ denote the intial 1-$g_0$, and low-g states, and the subscripts ${gl}$, ${sg}$, and ${ls}$ denote the gas-liquid, solid-gas and liquid-solid interfaces repectively. The motive force is due to the inertia of internal flows which occur as the interface deflects under the suddenely lessened $\mathbb{B}\mbox{o}$. The characteristic time scale of the rolling up scales as $t_j \sim (\rho V_d/\sigma)^{1/2}$\cite{attari_puddle_2016}, which resembles the contact time, $\tau \approx 2.6(\rho R^3_d/\sigma)^{1/2}$ presented by Richard \emph{et al.} in 2002 for the related problem of drops impacting hydrophobic surfaces in 1-$g_0$\cite{richard_surface_2002}. For drops with radial symmetry and sufficiently high initial $\mathbb{B}\mbox{o}$, inward radial capillary waves coalesce at the axis, leading to geysering and creation of satellite droplets by the Rayleigh-Plateau breakup of the geyser. In the case of smaller jumping drops the capillary waves are gradually damped by viscous forces. A time-lapsed composite photograph of spontaneous drop jump can be seen in Figure \ref{fig:nobounce}.
\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{drop_jump}
\caption{The trajectory of a 0.2 mL spontaneous drop jump is captured in a composite image over a period $\sim 0.67$ s at 120 fps. \label{fig:nobounce}}
\end{figure}

@@

\section{Drop Electro-Bounce}
During the `rolling-up' of drops under ideal conditions, the spontaneous jump phenomenon is governed by a balance of inertia and surface tension forces, and once aloft the drop motion is nominally in a regime of pure drag. However, other forces can again come into play.

We have observed jumped drops to occasionally decelerate and return to the superhydrophobic surface, rebounding multiple times in the fashion of rigid bodies bouncing under 1-$g_0$. The forces at work in such situations are presumably electrostatic in origin. Though this was an unintended effect during our work on spontaneous drop jump, the phenomenon is interesting itself. There is a certain irony to this typical failure of low-gravity intuition, where relatively tiny electrostatic forces have become suddenly large in the absence of gravity leading to unintentional experimental results, being as we are experienced low-gravity researchers who perhaps should know better. A time-lapsed composite image showing an example of the phenomenon is shown in Figure \ref{fig:bounce}.
\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{bounce}
\caption{The trajectory of a 0.5 mL drop is captured in a composite image over a single bounce period ($\sim 1.25$ s). The surface potential of the superhydrophobic dielectric in this example is $\varphi_s = 1.25 \pm 0.41$ kV. \label{fig:bounce}}
\end{figure}

@@

There are several possible origins of the electric charge which are ultimately responsible for this phenonenon. It is well known that water acquires positive free charge when in contact with certain polymers, especially polytetrafluoroethylene (PTFE)\cite{langmuir_surface_1938}, through a process called contact charging. PTFE, on the other hand, tends to readily acquire negative charge by contact with water. The superhydrophobic surfaces used in the spontanous drop jump experiments have thin (nanometric) PTFE coatings, and we observe that it is extremely easy to produce significant surface potentials, $\varphi_s \sim$ 100-500 V, by simply flowing streams of distilled water over them. A study of this water on PTFE contact charging phenomenon was conducted by Yatsuzuka \emph{et al.}\cite{yatsuzuka_electrification_1994}, who suggest that this process results from formation of an electrical double layer driven by selective adsorption of ($\mbox{OH}^-$) ions at the polymer surface; other recent work strongly supports this hypothesis\cite{beattie_intrinsic_2006}\cite{strazdaite_water_2015}. Yatsuzuka \emph{et al.} also found that the specific charge on the drops in contact with the surface depends on their both surface velocity and conductivity. The contact charging mechanism is the most likely accidental source of charge on the suphydrophobic surfaces used in our work. Given the large roughness, or the ratio of projected to actual surface area, of the superhydrophobic surfaces used in the experiment, and given that the drops are initially in a Cassie-Baxter state, a somewhat electrically resistive air layer is maintained that reduces grounding of the drops despite the large potential difference between them and and the surface charges. If this electrostatic potential is greater than the release of surface energy under the sudden change in $\mathbb{B}\mbox{o}$, in the form of the drop kinetic energy, we expect that the drop will return to an equilibrium state on the charged surface which minimizes the potential.

The source of the net free charge on the drops is another issue. The drop charge could of course also be due to the contact charging mechanism mentioned previously. For instance, in a 1996 paper, NASA flight engineer Don Pettit discusses the problem of low-gravity flow induced charging of liquids, resulting ultimately from contact charging phenomenon\cite{pettit_donald_flow_????}. Another mechanism for the drop charge is field-induced charging. Field-induced charging occurs due to physical breakup of a conductor having a field-induced dipole (e.g a physical seperation of charge). In our work this might occur when a drop is deposited on the charged surface by a grounded syringe. The metal syringe needle tip, and the liquid in the syringe itself are essentially a ground connection which is broken when the syringe is suddenly removed in the presence of an external electric field. Field-induced charging is at work in the famous Kelvin thunderstorm, and is applied in inkjet, and electrospay technologies, where in each case the breakup is by the Rayleigh-Plateau instability. Notably, in Pettit's aforementioned discussion of contact charging of liquids in low-g, he remarks on accidental electrostatic `hula-ing' of silicone oil drops when ejected from a syringe in the vicinity of a highly-charged polymer surface during an experiment conducted aboard STS-5 by mission specialist Joseph Allen\cite{pettit_donald_flow_????}. Depending on the (highly-variable) electrical conductivity of the silicone oil, and the material of the syringe used in the experiment, the charge could arise just as easily by field-induction as by contact charging. Relatedly, in a series of informal, and somewhat whimsical experiments, Pettit himself electrostatically orbited small water drops around a triboelectrically charged PTFE knitting needle while aboard ISS during expedition 30/31\cite{stevenson_electrostatic_2015}. Again, in this case, the drop charge is likely field-induced.

\section{Applications}
This study of the drop electro-bounce phenomenon may have results which can be extended to useful applications in a space environment. External surfaces of spacecraft tend to become charged with time due to the space environment itself. Surface charging largely occurs due to low energy electrons (3-50 keV) which do not penetrate the surface of the spacecraft external stucture\cite{czepiela_charging_1997}. The ultimate sources of these charges are trapped charged particles of the van Allen radiation belts, galactic cosmic rays, and the solar wind. This deposited charge accumulates and can lead to significant potential differences between different parts of a spacecraft, sometime leading to breakdown and discharge, called Paschen discharge. Deep dielectric charging occurs when higher energy charged particles penetrate the surface of a dielectric material, which can also lead to large potential differences if the dielectric leakage is lower than the external charging rate. In a fluid mechanical context, this accumulated charge could be problematic during any multiphase venting process, as might be expected to occur during autonomous satellite servicing refueling operations, or during boil-off venting of cryogens stored over long periods in a space environment (as in many crewed Mars mission scenarios, or in cryogenic propellant depot operations). Such venting, if occurring in a region of strong field, will also tend to produce a stream of charged drops, which will be decelerated by the field, causing their eventual return, impacting and potentially contaminating the spacecraft surfaces.

As previously noted, robust phase separation is critical to high reliability multiphase systems used in thermal control and life support. Active (but solid-state) phase seperation for life support multiphase flows, espcially high void-fraction droplet-disperse flows is a possible application of this work. Such flows are encountered in dish washing, laundry, waste solids drying, food processing, the Sabatier CO$_2$ reaction, and possibly in vapor-compression cycle condensers. Phase seperation for other disperse droplet flows include electrostatic droplet seperators for high-efficiency Rankine-cycle turbines, which require a droplet-free vapor phase entering the turbine, but where conventional superheat approaches comes with severe mass penalties\cite{unterberg_zero_1962}, or more speculatively, in high temperature liquid droplet radiators with electrostatic collection\cite{white_liquid_1987}. Removal of satellite droplets produced during pippetting in wet-lab research outside of a glovebox environment aboard ISS has also been recently suggested as a application of the work [private communication, NASA JSC, E. Unger, 2017]. Drops can become spontaneously charged by contact with standard micropipette tips\cite{choi_spontaneous_2013}, and this free charge can possibly be leveraged for the purposes of phase separation in low-gravity.

\section{Overview of this Thesis}
In this thesis we develop, and partly validate a simple model to aid in design of future electrostatic disperse drop phase separators. To this end the work encompasses:
\begin{enumerate}
\item A mathematical model of the drop electro-bounce, its scale analysis, asymptotic estimates for drop apoapses and times-of-flight.
\item The results of controlled experiments, where the effect of key independent variables drop volume $V_d$, and dielectric surface charge density $\sigma$, on drop trajectories is tested. 
\end{enumerate}
